---
title: "R Notebook"
output: 
  html_notebook: 
    toc: yes
editor_options: 
  chunk_output_type: inline
---

Compare STR calls using different software (NanoRepeat, EPI2ME-straglr). Also compare STR calls for individuals in a trio to check for Mendelian inheritance.

```{r}

library(tidyverse)
library(paletteer)
library(cowplot)
library(paletteer)

```

```{r}

setwd("/nesi/nobackup/uoa04099/benchmarking_STRs/")
mydir = "/nesi/nobackup/uoa04099/benchmarking_STRs/"
id = "HG005"

fname_nano = paste(mydir, id, "_nanorepeat_summarised.tsv", sep="")
fname_straglr = paste(mydir, id, "_straglr_summarised.tsv", sep="")
outfname = paste(mydir, "compare_outputs/", id, "_straglr_nano_compare.tsv", sep="")

in_nano = read.table(fname_nano,header=T, sep="\t")
in_straglr = read.table(fname_straglr,header=T, sep="\t")

```

##### Inner-join of straglr and nanorepeat results
Output is table with one locus per line, showing allele sizes for nanorepeat and straglr

Note that the result is a list of those loci that appear in both lists (nano and straglr) A bunch of loci were lost in the pipeline (62 input: 48 straglr, 47 nano)

```{r}
straglr = in_straglr %>% select(chr, start_pos, end_pos, gene, status, min_path, allele1, allele2, max_normal, min_path) %>% rowwise() %>% mutate(strg_a1=min(allele1, allele2)) %>% mutate(strg_a2=max(allele1, allele2))

straglr$start_pos = straglr$start_pos-1
nano = in_nano
names(nano)[5:7] = c("nano_num_alleles", "nano_a1", "nano_a2")
out = inner_join(straglr, nano, join_by(start_pos==start))
out = out %>% select(gene, motif, chr.x, start_pos, end_pos, nano_a1, nano_a2, strg_a1, strg_a2, nano_num_alleles, max_normal, min_path, status)
names(out)[3]="chr"

write_tsv(out, outfname)


```


Read in files for all individuals genotyped, calculate difference in allele size for nano and straglr

```{r}

list_of_files <- list.files(path = "/nesi/nobackup/uoa04099/benchmarking_STRs/compare_outputs", recursive = TRUE, pattern = "\\.tsv$", full.names = TRUE)

df <- readr::read_tsv(list_of_files, id = "file_name")

pattern1="/nesi/nobackup/uoa04099/benchmarking_STRs/compare_outputs/"
pattern2= "_straglr_nano_compare.tsv"

df2 = df %>% mutate(a1_diff = nano_a1 - strg_a1, a2_diff = nano_a2 - strg_a2, id_temp = gsub(pattern1, "", file_name)) %>% mutate(id = gsub(pattern2, "", id_temp))


```

Make plot

```{r}


# Reorder data using average? Learn more about reordering in chart #267
df3 <- df2 %>% 
  rowwise() %>%
  group_by(gene) %>%
  mutate( mymean = mean(c(a1_diff,a2_diff) )) %>% 
  arrange(mymean) %>% 
  mutate(gene=factor(gene))


# Plot
p1 = ggplot(df3) +
  geom_segment(aes(x=fct_rev(gene), xend=gene, y=a1_diff, yend=a2_diff),color="grey") +
  geom_hline(yintercept=0,lty=3) +
  geom_point(aes(x=fct_rev(gene), y=a1_diff,col=id), alpha=0.5, size=3 ) +
  geom_point(aes(x=fct_rev(gene), y=a2_diff, col=id), alpha=0.6, size=3 ) +
  #scale_color_paletteer_d("lisa::AndyWarhol_2") +
  scale_color_paletteer_d("rcartocolor::Antique")+
  coord_flip()+
  theme_cowplot() +
  theme(legend.position = "right") +
  xlab("") +
  ylab("Allele size difference (nano - straglr)")

save_plot('/nesi/nobackup/uoa04099/benchmarking_STRs/nano_straglr_compare_5.pdf', p1, base_height=10, base_width=7)

```


```{r}

p2temp = ggplot(df3) +
  geom_segment(aes(x=fct_rev(gene), xend=gene, y=a1_diff, yend=a2_diff),color="grey") +
  geom_hline(yintercept=0,lty=3) +
  geom_point(aes(x=fct_rev(gene), y=nano_a1,col=id), alpha=0.5, size=3 ) +
  geom_point(aes(x=fct_rev(gene), y=nano_a2, col=id), alpha=0.5, size=3 ) +
  geom_point(aes(x=fct_rev(gene), y=strg_a1,col=id), alpha=0.5, size=3 ) +
  geom_point(aes(x=fct_rev(gene), y=strg_a2, col=id), alpha=0.5, size=3 ) +
  geom_point(aes(x=fct_rev(gene), y=min_path), pch=15, col="firebrick3", alpha=0.3, size=3 ) +
  geom_point(aes(x=fct_rev(gene), y=max_normal), pch=15, col="dodgerblue3", alpha=0.3, size=3 ) +
  scale_color_paletteer_d("lisa::AndyWarhol_2") +
  coord_flip()+
  theme_cowplot() +
  theme(legend.position = "right") +
  xlab("") +
  ylab("Allele size ")

p2 = p2temp + annotate("text", x=24, 540, label="pathogenic_min", col="firebrick3") + 
  annotate("segment", x=24, xend=24, y=420, yend=320, col="firebrick3", arrow = arrow(type = "closed", length = unit(0.02, "npc")), alpha=0.6)+
  annotate("text", x=21, 470, label="normal_max", col="dodgerblue3") + 
  annotate("segment", x=21, xend=23.5, y=390, yend=260, col="dodgerblue3", arrow = arrow(type = "closed", length = unit(0.02, "npc")), alpha=0.6)
save_plot('/nesi/nobackup/uoa04099/benchmarking_STRs/str_allele_sizes.pdf', p2, base_height=10, base_width=7)


p3 = ggplot(df3) +
  geom_segment(aes(x=fct_rev(gene), xend=gene, y=a1_diff, yend=a2_diff),color="grey") +
  geom_hline(yintercept=0,lty=3) +
  geom_point(aes(x=fct_rev(gene), y=nano_a1,col=id), alpha=0.5, size=3 ) +
  geom_point(aes(x=fct_rev(gene), y=nano_a2, col=id), alpha=0.5, size=3 ) +
  geom_point(aes(x=fct_rev(gene), y=strg_a1,col=id), alpha=0.5, size=3 ) +
  geom_point(aes(x=fct_rev(gene), y=strg_a2, col=id), alpha=0.5, size=3 ) +
  geom_point(aes(x=fct_rev(gene), y=min_path), pch=15, col="firebrick3", alpha=0.3, size=3 ) +
  geom_point(aes(x=fct_rev(gene), y=max_normal), pch=15, col="dodgerblue3", alpha=0.3, size=3 ) +
  scale_color_paletteer_d("lisa::AndyWarhol_2") +
  coord_flip()+
  theme_cowplot() +
  theme(legend.position = "right") +
  xlab("") +
  ylab("Allele size ") + ylim(0,200) 

save_plot('/nesi/nobackup/uoa04099/benchmarking_STRs/str_allele_sizes_trunc2.pdf', p3, base_height=10, base_width=7)
p3

```


Make a plot that shows allele size and whether it falls into pathogenic or normal range
This requires the STRchive database file, which has columns for min and max pathogenic and max normal.
(I had just the pathogenic threshold and normal max, so needed to get this additional information from STRchive.)

From the STRchive github (edited)
# Load libraries, set theme & options
```{r}

library('plotly')
library('ggplot2')
library('cowplot')
theme_set(theme_cowplot())
options(stringsAsFactors = FALSE)

```

# Read in STRchive database files
```{r}

setwd("/nesi/nobackup/uoa04099/scripts/str-scripts/")

disease.loci = read.csv('/nesi/nobackup/uoa04099/benchmarking_STRs/STRchive_data/STRchive-database.csv')

fname_allele_size = '/nesi/nobackup/uoa04099/benchmarking_STRs/STRchive_data/STRchive_allele_size'
fname_age_onset = '/nesi/nobackup/uoa04099/benchmarking_STRs/STRchive_data/STRchive_age_onset.html'

```

# From STRchive plot.Rmd: switch between units (from repeat unit length to bp)
```{r}
disease.loci$norm_min_bp = disease.loci$normal_min * disease.loci$repeatunitlen
disease.loci$norm_max_bp = disease.loci$normal_max * disease.loci$repeatunitlen
disease.loci$int_min_bp = disease.loci$intermediate_min * disease.loci$repeatunitlen
disease.loci$int_max_bp = disease.loci$intermediate_max * disease.loci$repeatunitlen
disease.loci$path_min_bp = disease.loci$pathogenic_min * disease.loci$repeatunitlen
disease.loci$path_max_bp = disease.loci$pathogenic_max * disease.loci$repeatunitlen

disease.loci$norm_min_bp[disease.loci$norm_min_bp == 0] = NA
disease.loci$norm_max_bp[disease.loci$norm_max_bp == 0] = NA

disease.loci$disease_id = reorder(disease.loci$disease_id, -disease.loci$path_min_bp)

disease.loci = subset(disease.loci, !grepl("conflicting evidence", disease.loci$details))
```

# Merge the STRchive data into my data and plot
# Plot nanoRepeat and straglr-epi2me together
```{r}
cols=paletteer_d("lisa::AndyWarhol_2", n = 5)

#cols = c(HG002="#D32934FF", HG003="#2F191BFF", HG004="#2BAA92FF", Pathogenic="#D12E6CFF", Intermediate="#F4BCB9FF", Benign="dodgerblue3")

#cols_indiv = c(HG002="#D32934FF", HG003="#2F191BFF", HG004="#2BAA92FF")
cols_indiv = c(HG005="#D32934FF", HG006="#2F191BFF", HG007="#2BAA92FF")
cols_effect = c(Pathogenic="firebrick3", Intermediate="#F4BCB9FF", Benign="dodgerblue3")

df4 = left_join(df3, disease.loci, join_by(gene==gene))
names(df4)[18]="id"

p4 = ggplot(df4) +
  #geom_segment(aes(x=fct_rev(gene), xend=gene, y=a1_diff, yend=a2_diff),color="grey") +
  geom_hline(yintercept=0,lty=3) +
  geom_linerange(aes(x=fct_rev(gene),ymin=path_min_bp, ymax=path_max_bp), color=cols_effect['Pathogenic'],lwd=3,alpha=0.1) +
  geom_linerange(aes(x=fct_rev(gene),ymin = int_min_bp, ymax=int_max_bp), color=cols_effect['Intermediate'],lwd=3,alpha=0.1) +
  geom_linerange(aes(x=fct_rev(gene),ymin = norm_min_bp, ymax = norm_max_bp), color=cols_effect['Benign'],lwd=3,alpha=0.1) +
  scale_color_manual(values=cols_effect)+
  geom_point(aes(x=fct_rev(gene), y=nano_a1,col=id), alpha=0.5, size=3 ) +
  geom_point(aes(x=fct_rev(gene), y=nano_a2, col=id), alpha=0.5, size=3 ) +
  geom_point(aes(x=fct_rev(gene), y=strg_a1,col=id), alpha=0.5, size=3 ) +
  geom_point(aes(x=fct_rev(gene), y=strg_a2, col=id), alpha=0.5, size=3 ) +
  scale_colour_manual(values=cols_indiv) + coord_flip() + 
  theme_cowplot() + theme(legend.position = "right") +
  xlab("") + ylab("Allele size ") + ylim(0,200) 

save_plot('/nesi/nobackup/uoa04099/benchmarking_STRs/HG567_comparison.pdf', p4, base_height=10, base_width=7)
p4


```

# TRIO PLOTS -- nanoRepeat
```{r}
cols=paletteer_d("lisa::AndyWarhol_2", n = 5)
cols_effect = c(Pathogenic="firebrick3", Intermediate="#F4BCB9FF", Benign="dodgerblue3")

#cols_indiv = c(HG002="#D32934FF", HG003="#2F191BFF", HG004="#2BAA92FF")
#shapes_indiv = c(HG004=3,HG003=4,HG002=1)

cols_indiv = c(HG005="#D32934FF", HG006="#2F191BFF", HG0074="#2BAA92FF")
shapes_indiv = c(HG007=3,HG006=4,HG005=1)

df4 = left_join(df3, disease.loci, join_by(gene==gene))
names(df4)[18]="id"

df5 = df4 %>% filter(id!='KOLF2')

p5 = ggplot(df5) +
  geom_hline(yintercept=0,lty=3) +
  geom_linerange(data=disease.loci, aes(x=fct_rev(gene),ymin=path_min_bp, ymax=path_max_bp), color=cols_effect['Pathogenic'],lwd=3,alpha=0.3) +
  geom_linerange(data=disease.loci, aes(x=fct_rev(gene),ymin = int_min_bp, ymax=int_max_bp), color=cols_effect['Intermediate'],lwd=3,alpha=0.3) +
  geom_linerange(data=disease.loci, aes(x=fct_rev(gene),ymin = norm_min_bp, ymax = norm_max_bp), color=cols_effect['Benign'],lwd=3,alpha=0.3) +
  scale_color_manual(values=cols_effect) + coord_flip() + 
  geom_point(aes(x=fct_rev(gene), y=nano_a1,col=id, pch=id), alpha=0.5, size=3 ) +
  geom_point(aes(x=fct_rev(gene), y=nano_a2, col=id, pch=id), alpha=0.5, size=3 ) +
  scale_shape_manual(values=shapes_indiv) +
  scale_colour_manual(values=cols_indiv) + coord_flip() + 
  theme_cowplot() + theme(legend.position = "right") +
  xlab("") + ylab("Allele size ") + ylim(0,200) + ggtitle("nanoRepeat Allele Sizes")

save_plot('/nesi/nobackup/uoa04099/benchmarking_STRs/HG567_comparison_nanorepeat.pdf', p5, base_height=10, base_width=7)
p5


```

# TRIO PLOTS -- EPI2ME
```{r}
cols=paletteer_d("lisa::AndyWarhol_2", n = 5)

#cols = c(HG002="#D32934FF", HG003="#2F191BFF", HG004="#2BAA92FF", Pathogenic="#D12E6CFF", Intermediate="#F4BCB9FF", Benign="dodgerblue3")

cols_indiv = c(HG002="#D32934FF", HG003="#2F191BFF", HG004="#2BAA92FF")
cols_effect = c(Pathogenic="firebrick3", Intermediate="#F4BCB9FF", Benign="dodgerblue3")
shapes_indiv = c(HG004=3,HG003=4,HG002=1) # circle=1, plus=3, cross=4 [circle should be offspring]

df4 = left_join(df3, disease.loci, join_by(gene==gene))
names(df4)[18]="id"

df5 = df4 %>% filter(id!='KOLF2')

p5 = ggplot(df5) +
  geom_hline(yintercept=0,lty=3) +
  geom_linerange(data=disease.loci, aes(x=fct_rev(gene),ymin=path_min_bp, ymax=path_max_bp), color=cols_effect['Pathogenic'],lwd=3,alpha=0.3) +
  geom_linerange(data=disease.loci, aes(x=fct_rev(gene),ymin = int_min_bp, ymax=int_max_bp), color=cols_effect['Intermediate'],lwd=3,alpha=0.3) +
  geom_linerange(data=disease.loci, aes(x=fct_rev(gene),ymin = norm_min_bp, ymax = norm_max_bp), color=cols_effect['Benign'],lwd=3,alpha=0.3) +
  scale_color_manual(values=cols_effect) + coord_flip() + 
  geom_point(aes(x=fct_rev(gene), y=strg_a1,col=id, pch=id), alpha=0.5, size=3 ) +
  geom_point(aes(x=fct_rev(gene), y=strg_a2, col=id, pch=id), alpha=0.5, size=3 ) +
  scale_shape_manual(values=shapes_indiv) +
  scale_colour_manual(values=cols_indiv) + coord_flip() + 
  theme_cowplot() + theme(legend.position = "right") +
  xlab("") + ylab("Allele size ") + ylim(0,1000) + ggtitle("EPI2ME Allele Sizes")

save_plot('/nesi/nobackup/uoa04099/benchmarking_STRs/HG567_comparison_epi2me.pdf', p5, base_height=10, base_width=7)
p5


```

 
# TRIO PLOTS -- nanoRepeat, rescale
```{r}

curr_trio=c('HG005', 'HG006', 'HG007')
plot_title = paste('Nanorepeat: ', paste(curr_trio[1], curr_trio[2], curr_trio[3], sep=','))

cols=paletteer_d("lisa::AndyWarhol_2", n = 5)
cols_effect = c(Pathogenic="firebrick3", Intermediate="#F4BCB9FF", Benign="dodgerblue3")

cols_indiv = c(HG005="#D32934FF", HG006="#2F191BFF", HG0074="#2BAA92FF")

shapes_indiv = c(HG005=1, HG006=4, HG007=3) # shape1=circle, shape3=plus, shape4=cross 
                                            # HG005=son(circle), HG006=father(cross),HG007=mother(plus)

df4 = left_join(df3, disease.loci, join_by(gene==gene))
names(df4)[18]="id"

df5 = df4 %>% filter(id %in% curr_trio) # select just our current trio
df5 = df5 %>% mutate(nano_a1_dev = nano_a1-max_normal) %>% mutate(nano_a2_dev = nano_a2-max_normal)

p5 = ggplot(df5) +
  geom_hline(yintercept=0,lty=3) +
  scale_color_manual(values=cols_effect) + coord_flip() + 
  geom_point(aes(x=fct_rev(gene), y=nano_a1_dev,col=id, pch=id), alpha=0.5, size=3 ) +
  geom_point(aes(x=fct_rev(gene), y=nano_a2_dev, col=id, pch=id), alpha=0.5, size=3 ) +
  scale_shape_manual(values=shapes_indiv, labels=c("offspring", "father", "mother")) +
  scale_colour_manual(values=cols_indiv) + guides(color = "none") + coord_flip() + 
  theme_cowplot() + theme(legend.position = "right") +
  xlab("") + ylab("Allele Size Deviation from Max Normal") + ylim(-200,200) + ggtitle(plot_title)

save_plot('/nesi/nobackup/uoa04099/benchmarking_STRs/HG567_nanorepeat_maxnormal.pdf', p5, base_height=10, base_width=7)
p5


```
